---
title: "Power curves using simr and open datasets"
output: html_notebook
---

Code for MA project investigating statistical power for small effect sizes. Full thesis text can be found here:  https://dspace.library.uvic.ca/items/535f7ac3-1783-4ca8-bdf3-0ca6487baa87

Perea et al. (2015) data:

- Masked priming lexial decision task
- 40 participants ("SUBJECT")
- 120 stimuli items ("ITEM")
- experimental conditions: repetition vs. unrelated priming ("REPETITION")
- also includes case alternation but we won't be using that
- 37ms priming effect between repetition conditions
- already cleaned for outliers <250 ms and >1500 ms
- full citation: Perea, M., Vergara-Martínez, M., & Gomez, P. (2015). Resolving the locus of cAsE aLtErNaTiOn effects in visual word recognition: Evidence from masked priming. Cognition, 142, 39–43. https://doi.org/10.1016/j.cognition.2015.05.007
- methodology from Brysbaert, M., & Stevens, M. (2018). Power Analysis and Effect Size in Mixed Effects Models: A Tutorial. Journal of Cognition, 1(1), 1–20. https://doi.org/10.5334/joc.10
- citation for simr: Green, P., & MacLeod, C. J. (2016). simr: An R package for power analysis of generalized linear mixed models by simulation. Methods in Ecology and Evolution, 7(4), 493–498. https://doi.org/10.1111/2041-210X.12504



# Preparing the dataset

```{r}
# loading the appropriate libraries
library(tidyverse)
library(lme4)
library(optimx)
library(afex)
library(simr)


# loading and cleaning data
perea37 <- read.csv("data/perea_for_analysis.csv")
perea37 <- na.omit(perea37)
head(perea37)


# taking a look at the distribution of the data
unrel_only <- subset(perea37, REPETITION=="unrelated")
hist(unrel_only$RT, 50)
mean(unrel_only$RT)

rep_only <- subset(perea37, REPETITION=="repeated")
hist(rep_only$RT, 50)
mean(rep_only$RT)
```


# Functions for finding and setting priming effect

These functions find and alter the priming effect, aka the difference in mean RT between the related and unrelated conditions. The second function adds a constant to each RT in the "related" condition to alter the priming effect. Since this is a linear transformation, the shape of the distribution is not affected and the relationships between each data point are preserved. Linear transformations like these are used generally uncontroversially throughout social sciences methodologies.

```{r}
# function for finding the priming effect (will come in handy later)
priming_effect <- function(x) {
  unrel_pe <- subset(x, REPETITION=="unrelated")
  rel_pe <- subset(x, REPETITION=="repeated")
  return(mean(unrel_pe$RT) - mean(rel_pe$RT))
}

# testing priming_effect, should output 36.95764
priming_effect(perea37)


# function for creating specific priming effect
# x = dataset to modify (always perea37), y = intended priming effect
make_effect <- function(x, y) {
  pe <- priming_effect(x)
  toadd <- pe - y
  unrel_me <- subset(x, REPETITION=="unrelated")
  rel_me <- subset(x, REPETITION=="repeated")
  if(toadd > 0)
    rel_me$RT = rel_me$RT+toadd
  if(toadd < 0)
    rel_me$RT = rel_me$RT-abs(toadd)
  x <-rbind(rel_me,unrel_me)
  return(x)
}

# testing make_effect, should output 15
perea15 <- make_effect(perea37, 15)
priming_effect(perea15)

# double-checking make_effect, manually calculating priming effect in the dataset
temp_unrel_only <- subset(perea15, REPETITION=="unrelated")
temp_rep_only <- subset(perea15, REPETITION=="repeated")
mean(temp_unrel_only$RT)-mean(temp_rep_only$RT)
```

# Power curve for 37ms effect along ITEM and SUBJECT

Testing the technique by replicating the results from Brysbaert & Stevens (2018). This section uses the powerCurve function from simr to assess power for increasing numbers of items (keeping number of participants constant) or increasing number of participants (keeping number of items constant). nsim in the powerCurve function is used to set the number of simulated datasets used at each level.

```{r}
# add invRT because Brysbaert & Stevens (2018) did it
perea37$invRT = -1000/perea37$RT

# checking priming effect
priming_effect(perea37)

# setting model for power curve
# optimx optimizer lets full random effects structure converge
fit37 <- lmer(invRT ~ REPETITION + (1|ITEM) + (1 + REPETITION|SUBJECT),
            data=perea37,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

summary(fit37)

# curve 1: 37ms effect along ITEM
pc1 <- powerCurve(fit37, along="ITEM", nsim=50)
plot(pc1) + abline(v=39) +
  title(main = "Power Curve for 37ms priming effect along ITEM")

# curve 2: 37ms effect along SUBJECT
pc2 <- powerCurve(fit37, along="SUBJECT", nsim=50)
plot(pc2) + abline(v=6.5) +
  title(main = "Power Curve for 37ms priming effect along SUBJECT")
```


# Power curve for 11.38ms effect along ITEM and SUBJECT

This section applies the technique from above to an 11.38ms effect to investigate the necessary numbers of items and participants to detect an effect so small, compared to the 37ms priming effect used in the tests above.


## Preparing the dataset and fitting a glm

```{r}
# adjust priming effect using make_effect
perea11 <- make_effect(perea37, 11.38)

# checking output of make_effect
priming_effect(perea11)

temp_unrel_only <- subset(perea11, REPETITION=="unrelated")
hist(temp_unrel_only$RT, 50)
mean(temp_unrel_only$RT)

temp_rep_only <- subset(perea11, REPETITION=="repeated")
hist(temp_rep_only$RT, 50)
mean(temp_rep_only$RT)


# fitting a glm to the dataset
# only random effects by intercept are included because including all random slopes and random slopes
# by participant caused convergence errors
# random slopes by item would not make sense, as the items do not cross the repetition condition
# see "Things that did not work" at the end for more details
fit <- glmer(RT ~ REPETITION + (1|ITEM) + (1|SUBJECT),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

summary(fit)

#testing model fit by plotting residuals
res <- resid(fit)
plot(fitted(fit), res) + abline(h=0)
plot(res)
qqnorm(res) + qqline(res)

# testing model fit by AIC, model with random intercepts results in lowest AIC even with added parameters
fit_participant_only<- glmer(RT ~ REPETITION + (1|SUBJECT),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

fit_item_only <- glmer(RT ~ REPETITION + (1|ITEM),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

fit_no_random_effects <- lm(RT ~ REPETITION,
            data=perea11)

anova(fit,fit_participant_only,fit_item_only,fit_no_random_effects)
```


## Power curve for 40 participants

More items than were available in the original dataset were necessary to reach 80% power at the given levels, so simulation was used again to create a dataset with 200 items, based on the glm fitted in the previous chunk. The number of simulations at each level was also increased to 1000, compared to the 50 that were used in Brysbaert & Stevens (2018), to reduce error.

```{r}
# simulating more items
more_items <- extend(fit, along="ITEM", n=200)

# power curve for 11.38ms effect along ITEM, including extending number of items to 200
# note that nsim = 1000, compared to 50 in the Brysbaert & Stevens (2018) protocol
pc_moreitems <- powerCurve(more_items, along="ITEM", breaks=c(100,120,140,160,180,200), nsim=1000)
plot(pc_moreitems) +abline(v=145) +
  title(main = "Estimated power for an 11.38ms priming effect
        with 40 participants and 100-200 items")

# spot-testing power for 145 items, 40 participants, given results of powerCurve
item145 <- extend(fit, along="ITEM", n=145)
power_item145 <- powerSim(item145)
```


## Power curve for 120 items

More participants than were available in the original dataset were necessary to reach 80% power at the given levels, so simulation was used again to create a dataset with 80 participants, based on the glm fitted in the previous chunk. The number of simulations at each level was again increased to 1000.

```{r}
# simulating more participants 
more_participants <- extend(fit, along="SUBJECT", n=80)

# power curve for 11.38ms effect along PARTICIPANT, including extending number of items to 200
# note that nsim = 1000, compared to 50 in the Brysbaert & Stevens (2018) protocol
more_participants <- powerCurve(more_participants, along="SUBJECT", breaks=c(30,35,40,45,50,55,60,65,70,75,80), nsim=1000)
plot(more_participants) + abline(v=45) +
  title(main = "Estimated power for an 11.38ms priming effect
        with 120 items and 30-80 participants")

# spot-testing power for 120 items, 47 participants, given results of powerCurve
participant47 <- extend(fit, along="SUBJECT", n=47)
power_participant47 <- powerSim(participant47)
```



## Things that did not work

Tracking all the things that didn't work in this project so that I don't try to do them again! The glms that resulted in convergence errors are noted here, as well as some ways of saving the output of powerCurve (saving as rdata and png)

```{r}
# all the models that I tried to fit that didn't work

# fitting all slopes result in convergence errors
fit_all_slopes <- glmer(stdRT ~ REPETITION + (1+REPETITION|ITEM) + (1+REPETITION|SUBJECT),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

summary(fit_all_slopes) #convergence errors

# fitting with random intercepts and slope by item did not result in a convergence error, but makes no sense
# REPETITION doesn't cross ITEM
fit_item <- glmer(stdRT ~ REPETITION + (1+REPETITION|ITEM) + (1|SUBJECT), # + (0+REPETITION|SUBJECT),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

summary(fit_item) #converged but makes no sense

# fitting with random intercepts and slope by participant resulted in convergence errors
fit_participant <- glmer(stdRT ~ REPETITION + (1|ITEM) + (1+REPETITION|SUBJECT), # + (0+REPETITION|SUBJECT),
            data=perea11,
            family=gaussian,
            control=lmerControl(optimizer = "optimx", calc.derivs = FALSE, optCtrl = list(method = "nlminb", starttests = FALSE, kkt =FALSE)))

summary(fit_participant) #convergence error

## SAVE AS RDATA FOR POWER CURVES DOESN'T WORK
save(pc3, file="pc_item_11ms.RData")
load("pc3") #ERROR: cannot open compressed file 'pc3', probable reason 'No such file or directory'Error in readChar(con, 5L, useBytes = TRUE) : cannot open the connection

## saving plot as PNG also didn't work?
png('pc_item_plot.png')
plot(pc3) + abline(v=94) +
  title(main = "Power Curve for 11.38ms priming effect along ITEM")
dev.off()

```